---
title: "Descriptive statistics"
format: 
  html: 
    code-fold: true
    embed-resources: true 
editor: visual
---

```{=html}
<style>
details > summary {
  padding: 4px;
  width: 400px;
  background-color: #eeeeee;
  border: none;
  box-shadow: 1px 1px 2px #bbbbbb;
  cursor: pointer;
}

details > p {
  background-color: #E0F8EC;
  padding: 4px;
  margin: 0;
  box-shadow: 1px 1px 2px #bbbbbb;
}
header h1 {
      font-size: 70px;
      font-weight: 600;
      font-palette: tomato;
    }
</style>
```

```{r, setup, echo=F, warning=FALSE,message=FALSE}
require("knitr")
library(readr)
library(DT)
library(kableExtra)
library(tidyverse)
clinical_data <- read_csv("Datasets/clinical_data.csv")
```
[\<\<\< Back to Index](./course_index.html)

**Descriptive statistics** summarize and describe the characteristics of a **population** or a **sample** of a population. There are several types of characteristics we can study to extract patterns, trends and the distribution of data from a dataset.

- **Data representation** : Allow to visualize how variables behave in our dataset.

- **Measure of central tendency** : Describe the center of the data (***mean***, ***median***, ***mode***).

- **Measure of variability** : Describe the dispersion of the data (***variance***, ***standard deviation***).

- **Measure of frequency distribution** : Describe the occurrence of data (***count***).

## Toy dataset : `clinical_data`

We are going to use a synthetic dataset containing 150 patients who went to the hospital due to symptoms corresponding to possible cardiovascular problems. For each patient, 8 observations (variables) were recorded. 

```{r, echo=F}
datatable(clinical_data, options = list(pageLength = 10))
```

## Types of variables

Before diving into the data, it is important to know them. A major point that will drive the analytical strategy is to know what variable(s) we have and what type they belong to.

Generally, variables are divided into two categories : **quantitative** and **qualitative**.

- **Quantitative** : Corresponds to a variable that can be counted/quantified
- **Qualitative** : Corresponds to a variable that can be categorized/ordered but **cannot be quantified**.

### Qualitative variables

- **Nominal** : Describes a name, a label or a category, one or more, without specific order.
- **Ordinal** : Defines an order relation (hierarchy) between the different categories.
- **Binary** : Nominal variables which have only two categories or level.

### Quantitative variables

- **Discrete** : Takes finite number of real and distinct values.
- **Continuous** : Takes non-finite number of values.

![](./Images/types_variables.drawio.svg){fig-align="center"}

### Types of variables in `clinical_data`

In our dataset `clinical_data`, we have several types of variables.

```{r, echo=F}
clinical_data |>
  head(10) |>
  kable() |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

<details>

<summary>Answer</summary>
::: {style="background-color:LemonChiffon;"}
- **Continuous** variables = *age*, *height*, *weight* and *glucose*.
- **Nominal** variables = *diagnosis*.
- **Ordinal** variables = *treatment_response*.
- **Nominal Binary** variables = *gender* and *smoking_status*.
:::
</details>

## Data representation

**Graphical representation** of certain variables can help to understand the different **characteristics** of the data.
This initial understanding of the data will help us formulate one or more **hypotheses** that can be verified or disproved later.

There are many ways to view data. We will show you a few that can usually be done using tools such as Excel, GraphPad, or programming languages such as R.

### Histogram

```{r, echo=F}
ggplot(clinical_data, aes(x = age)) +
  geom_histogram(binwidth = 2, color = "black", fill = "grey50") +
  theme_bw() +
  labs(x = "Age", y = "Frequency") +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
```

### Density plot

```{r, echo=F}
ggplot(clinical_data, aes(x = height)) +
  geom_density(aes(color=gender)) +
  theme_bw() +
  labs(x = "Height", y = "Frequency") +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
```

## Central tendency


## Variability / Dispersion

