---
title: "Exercices"
format: 
  html: 
    code-fold: true
    embed-resources: true 
editor: visual
---

```{=html}
<style>
details > summary {
  padding: 4px;
  width: 400px;
  background-color: #eeeeee;
  border: none;
  box-shadow: 1px 1px 2px #bbbbbb;
  cursor: pointer;
}

details > p {
  background-color: #E0F8EC;
  padding: 4px;
  margin: 0;
  box-shadow: 1px 1px 2px #bbbbbb;
}
header h1 {
      font-size: 70px;
      font-weight: 600;
      font-palette: tomato;
    }
</style>
```

```{r, setup, echo=F, warning=FALSE,message=FALSE}
require("knitr")
library(readr)
library(tidyverse)
library(FSA)
library(corrplot)
data_exercice <- read_csv("Datasets/data_exercice.csv")
```

# Objectives

The objective of this section is to train you in using the various statistical tests we have seen and interpreting them.
Several questions will be asked. It is up to you to understand them clearly and use the most appropriate statistical approach to answer them.

# About the data

The Pellegrin University Hospital in Bordeaux conducted a voluntary study on 150 patients to examine health factors such as blood sugar, cholesterol, and high blood pressure on a sample of the population of Bordeaux. The results are presented in the dataset called `data_exercise`. You'll need to explore this data a little. This is the first step to take before asking questions and diving into statistics.

  - **After loading the data, identify the types of variables in the dataset**
  
<details>

<summary>Types of variables</summary>

  - `sex` : **qualitative** *nominal*
  - `age` : **quantitative** *continuous*
  - `height_cm` : **quantitative** *continuous*
  - `weight_kg` : **quantitative** *continuous*
  - `bmi` : **quantitative** *continuous*
  - `glucose` : **quantitative** *continuous*
  - `cholesterol_total` : **quantitative** *continuous*
  - `bmi_category`: **qualitative** *ordinal*
  - `physical_activity`: **qualitative** *ordinal*
  - `hypertension` : **qualitative** *nominal binary*

</details>

  - **How to explore data with R?**

<details>

<summary>Explore data</summary>

```{r, eval=F}
summary(data_exercice)
```

```{r, echo=F}
summary(data_exercice)
```

</details>

# Questions

For each question, you have the analysis process described in the hidden section. You can consult it if you get stuck or if you want to check your reasoning!

### I -  **Is there a link between BMI and cholesterol levels?**

<details>

<summary>Response</summary>

The question asked seeks to determine whether there is a **link between two variables**. Both variables are **quantitative**. We will therefore perform a **correlation test** between `BMI` and `cholesterol` levels.

<details>

<summary>Visualisation of data</summary>

The first step is to look at the data and see how it behaves in a simple plot.

```{r, eval=F}
plot(data_exercice$bmi, data_exercice$cholesterol_total, 
     xlab = "BMI", 
     ylab = "Total cholesterol (g/L)")
```

```{r, echo=F}
plot(data_exercice$bmi, data_exercice$cholesterol_total, 
     xlab = "BMI", 
     ylab = "Total cholesterol (g/L)")
```

There is a slight trend where the two variables seem to follow each other, but to what extent?

</details>

<details>

<summary>Hypotheses</summary>

  - H0 = **The variables `BMI` and `cholesterol_total` are not related**
  - H1 = **The variables `BMI` and `cholesterol_total` are related**

</details>

<details>

<summary>Condition</summary>

The main parameter to check is the normality of the data. This can be done using the Shapiro test or by graphically visualizing the quantiles :

  - H0 = **The data follow a normal distribution**
  - H1 = **The data does not follow a normal distribution**
  
  
##### `cholesterol_total`
```{r, eval=F}
# Q-Q plot with qqnorm function for the cholesterol levels
qqnorm(data_exercice$cholesterol_total)
qqline(data_exercice$cholesterol_total)
```

```{r, echo=F}
qqnorm(data_exercice$cholesterol_total)
qqline(data_exercice$cholesterol_total)
```

```{r, eval=F}
# Shapiro test
shapiro.test(data_exercice$cholesterol_total)
```

```{r, echo=F}
shapiro.test(data_exercice$cholesterol_total)
```

##### `BMI`
```{r, eval=F}
# Q-Q plot with qqnorm function for the cholesterol levels
qqnorm(data_exercice$bmi)
qqline(data_exercice$bmi)
```

```{r, echo=F}
qqnorm(data_exercice$bmi)
qqline(data_exercice$bmi)
```

```{r, eval=F}
# Shapiro test
shapiro.test(data_exercice$bmi)
```

```{r, echo=F}
shapiro.test(data_exercice$bmi)
```

For both variables, we obtain **significant p-values**. We therefore **reject H0** in both cases, meaning that the data for these variables **do not follow a normal distribution**.

</details>

<details>

<summary>Test</summary>

Since the normality condition is not met, we use **Spearman's correlation**.

```{r, eval=F}
# Spearman correlation
cor.test(data_exercice$bmi, data_exercice$cholesterol_total, method = "spearman")
```

```{r, echo=F}
cor.test(data_exercice$bmi, data_exercice$cholesterol_total, method = "spearman")
```

The p-value obtained is **significant**, so we **reject H0** and consider that there is a **correlation between the two variables**. Furthermore, based on the r value obtained, we can say that the **correlation is weakly positive** (r around 0.2).

</details>

</details>

### II - **Does BMI differ depending on patients' physical activity levels?**

<details>

<summary>Response</summary>

We want to find out whether **BMI differs depending on patients' level of physical activity**. We will therefore compare patients by physical activity group based on their average BMI, **using an ANOVA-type analysis**.

<details>

<summary>Visualisation</summary>

```{r, eval=F}
ggplot(data = data_exercice, aes(x=physical_activity, y = bmi)) +
  geom_boxplot() + 
  labs(x="Levels of physical activity", y = "BMI") +
  theme_bw()
```

```{r, echo=F}
ggplot(data = data_exercice, aes(x=physical_activity, y = bmi)) +
  geom_boxplot() + 
  labs(x="Levels of physical activity", y = "BMI") +
  theme_bw()
```

It would appear that there is a difference in BMI between the physical activity groups based on this box plot representation, particularly with regard to the group that does little physical activity

</details>

<details>

<summary>Hypotheses</summary>

  - H0 = **The average `BMI` in each physical activity group is not different.**
  - H1 = **We have at least one physical activity group with a different average `BMI`.**

</details>

<details>

<summary>Condition</summary>

The main parameter to check is the normality of the data. This can be done using the Shapiro test or by graphically visualizing the quantiles :

  - H0 = **The data follow a normal distribution**
  - H1 = **The data does not follow a normal distribution**

##### `BMI`
```{r, eval=F}
# Q-Q plot with qqnorm function for the cholesterol levels
qqnorm(data_exercice$bmi)
qqline(data_exercice$bmi)
```

```{r, echo=F}
qqnorm(data_exercice$bmi)
qqline(data_exercice$bmi)
```

```{r, eval=F}
# Shapiro test
shapiro.test(data_exercice$bmi)
```

```{r, echo=F}
shapiro.test(data_exercice$bmi)
```

The BMI variable data **does not follow a normal distribution**. In fact, the **p-value is not significant**. 

</details>

<details>

<summary>Test</summary>

Since the normality is not respected, we use **Kruskal-Wallis test**.

```{r, eval=F}
# Kruskal-Wallis test
kruskal.test(data_exercice$bmi, data_exercice$physical_activity)
```

```{r, echo=F}
kruskal.test(data_exercice$bmi, data_exercice$physical_activity)
```

The p-value is below 0.05, this is **significant**. We can **reject H0** and consider that **at least one group has a different average BMI**.

</details>

<details>

<summary>Post-hoc</summary>

In a post-hoc test, we seek to compare the groups two by two, which amounts to positing the following hypotheses:

  - H0 : **There is no difference in average BMI between physical activity group A and B**.
  - H1 : **There is a difference in average BMI between physical activity group A and B**.

Since we performed a Kruskal-Wallis test, we will use **Dunn's post hoc test**.

```{r, eval=F}
# If not already installed, install the FSA package
install.packages("FSA")
# Activate the package
library(FSA)

# Dunn post-hoc
dunnTest(bmi ~ physical_activity, data = data_exercice, kw = F) #kw = F allow to not redo the Kruskal-Wallis test
```

```{r, echo=F}
dunnTest(bmi ~ physical_activity, data = data_exercice, kw = F)
```

All the p-values are **significant in all comparisons**. For every comparisons we can **reject H0**, and we can consider that there is a **difference in average BMI based on the physical activity**. We can notice with the Z-score that when the activity is low, the difference is higher with high practice than with moderate practice. We can say also that patients with low activity tend to have a higher BMI.

</details>

</details>

### III - **Is high blood pressure more common in women?**

<details>

<summary>Response</summary>

<details>

<summary>Visualisation</summary>

Firstly, we create a contingency table to see the repartition of patients with hypertension by sex.

```{r, eval=F}
# Contingency table between hypertension and sex
tab <- table(data_exercice$hypertension, data_exercice$sex)
tab
```

```{r, echo=F}
tab <- table(data_exercice$hypertension, data_exercice$sex)
tab
```

</details>

<details>

<summary>Hypotheses</summary>

  - H0 : **The proportion of hypertensive patient is the same in woman and man.**
  - H1 : **The proportion of hypertensive patient is different** 

</details>

<details>

<summary>Condition</summary>

The only condition here is to have a minimum **sample size of 5** to be respected for each class in the contingency table. This is the case here, so we can use **Chi-square homogeneity test**.

</details>

<details>

<summary>Test</summary>

```{r, eval=F}
# Contingency table between hypertension and sex
tab <- table(data_exercice$hypertension, data_exercice$sex)

# Chi-square homogeneity test
chisq.test(tab)
```

```{r, echo=F}
tab <- table(data_exercice$hypertension, data_exercice$sex)
chisq.test(tab)
```

The test gives a **significant p-value**. So we **reject H0**, and we consider that the **proportion of hypertensive patient is different between women and men**.
We need to go deeper by observing graphically this difference in terms of proportions.

```{r, eval=F}
# Observe proportions graphically
  ggplot(data_exercice, aes(x = sex, fill = hypertension)) + 
  geom_bar(position = "fill")+ 
  theme_bw() +
  labs(x = "", y = "Proportion")
```

```{r, echo=F}
ggplot(data_exercice, aes(x = sex, fill = hypertension)) + 
  geom_bar(position = "fill")+ 
  theme_bw() +
  labs(x = "", y = "Proportion")
```

What we see is that there is indeed a difference in proportions. But not the one we expected. In fact, **more men have high blood pressure than women**.

</details>

</details>

### IV - **Do patients with hypertension have higher blood sugar levels?**

<details>

<summary>Response</summary>

<details>

<summary>Visualisation</summary>

```{r, eval=F}
ggplot(data = data_exercice, aes(x=hypertension, y = glucose)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

```{r, echo=F}
ggplot(data = data_exercice, aes(x=hypertension, y = glucose)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

At first glance, these box plots **do not reveal a clear difference** in glucose levels between the two groups. This requires further investigation through **statistical analysis**.

</details>

<details>

<summary>Hypotheses</summary>

  - H0 : **The average glucose is not different in patients with hypertension and without hypertension**
  - H1 : **The average glucose is higher in patients with hypertension**
  
Here, the question is specific: **we want to know whether the value is higher** in one group compared to the other. Furthermore, the patients are **independent** of one another, so we should prioritize **tests for independent samples**.

</details>

<details>

<summary>Condition</summary>

First, we check whether the **data follows a normal distribution or not**.

  - H0 = **The data follow a normal distribution**
  - H1 = **The data does not follow a normal distribution**

```{r, eval=F}
# Q-Q plot with qqnorm function for the glucose levels
qqnorm(data_exercice$glucose)
qqline(data_exercice$glucose)
```

```{r, echo=F}
qqnorm(data_exercice$glucose)
qqline(data_exercice$glucose)
```

```{r, eval=F}
# Shapiro test
shapiro.test(data_exercice$glucose)
```

```{r, echo=F}
shapiro.test(data_exercice$glucose)
```

We obtain a **non-significant p-value**, and we can see many **outliers at the extremes on the QQ plot**. We **cannot reject H0**, and therefore we conclude that the data **do not follow a normal distribution**.
So we’ll start with a nonparametric test: **the Wilcoxon-Mann-Whitney U test**.

</details>

<details>

<summary>Test</summary>

```{r, eval=F}
# Wilcoxon-Mann-Whitnet U test
wilcox.test(data_exercice$glucose[data_exercice$hypertension == "Yes"],
            data_exercice$glucose[data_exercice$hypertension == "No"], 
            alternative = "greater",
            paired = FALSE)
```

```{r, echo=F}
# Wilcoxon-Mann-Whitnet U test
wilcox.test(data_exercice$glucose[data_exercice$hypertension == "Yes"],
            data_exercice$glucose[data_exercice$hypertension == "No"], 
            alternative = "greater",
            paired = FALSE) 
```

Here, the aim was to determine whether **patients with hypertension had higher glucose levels than others**. There was a sense of direction, hence the `alternative = “greater”`.

Since we specified that the **samples are independent**, we can set `paired = FALSE`, even though this argument defaults to `FALSE`.

The p-value obtained is **greater than 0.05**, so we **cannot reject H0**. This means that, statistically, **patients with hypertension do not have higher glucose levels than the others**.

</details>

</details>