---
title: "Practical stats"
format: 
  html: 
    code-fold: true
    embed-resources: true 
editor: visual
---

```{=html}
<style>
details > summary {
  padding: 4px;
  width: 400px;
  background-color: #eeeeee;
  border: none;
  box-shadow: 1px 1px 2px #bbbbbb;
  cursor: pointer;
}

details > p {
  background-color: #E0F8EC;
  padding: 4px;
  margin: 0;
  box-shadow: 1px 1px 2px #bbbbbb;
}
header h1 {
      font-size: 70px;
      font-weight: 600;
      font-palette: tomato;
    }
</style>
```

```{r, setup, echo=F, warning=FALSE,message=FALSE}
require("knitr")
library(readr)
library(tidyverse)
clinical_data <- read_csv("Datasets/clinical_data.csv")
```

# Objectives

The purpose of this section is to provide you with the basics needed to analyze biological data using simple statistics. The main objective is to provide you with the tools to :

-   Choose a statistical method appropriate for a biological question
-   Interpret the results correctly
-   Analyse data with R

# Discover and explore data : `clinical_data`

## Load the data

First step to do is to load our dataset called `clinical_data` in our R environment. You can follow the instruction below.

![](./Images/load_data.svg){fig-align="center"}

## Description of the data

The first thing to do before starting the analysis is to observe the dataset in order to define the structure and type of data. You can do it by hand, or you can use R and some functions already built in, without installing any packages.

### `head`

The first function you can use is `head`. It will returns the **first** (or **last** with `tail`) **rows** of your dataset. By default, **the 6 first (or last) rows** will be returned.

```{r head}
#| code-fold: false
head(clinical_data)
```

### `str`

The function `str` allows to display the **internal structure** of the data. Using this method, we can see the **dimensions** of our dataset, i.e., the number of rows and columns. Here we have **150 rows and 10 columns**. You can also see what **type** each column is, like **numerical** or **character**.

Thanks to this view, we can define more precisely what **type of data** we are dealing with. ([See the descriptive statistics part](./descriptive_stats.html))

```{r str}
#| code-fold: false
str(clinical_data)
```

### `summary`

The `summary` function is very useful because it provides **descriptive statistical results**, as seen in the previous section. For each column, we will find the **median**, the **mean**, and the **quartiles**, among other things.

```{r summary}
#| code-fold: false
summary(clinical_data)
```

# Study 1 : *Do patients with heart disease have a higher heart rate ?*

## First visualisation

We can start by making a simple observation using a **boxplot** of **heart rate** based on patient **diagnosis** :

```{r, eval=F}
ggplot(data = clinical_data, aes(x=diagnosis, y = heart_rate)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

```{r, echo=F}
ggplot(data = clinical_data, aes(x=diagnosis, y = heart_rate)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

Based on this initial visualization, we can see that there appears to be a **difference in heart rate among patients with heart disease**.

## Hypotheses : decision support

A statistical approach is based on reasoning with **two initial solutions** called working **hypotheses**. We find the **null hypothesis** (H0) and the **alternative hypothesis** (H1).

-   *Null hypothesis* (H0) : corresponds to the **initial assumption**; this hypothesis seeks to confirm that there are **no differences**, for example.
-   *Alternative hypothesis* (H1) : is the opposite, it suggests that **there is a difference or an effect**.

*In our case, what would be the hypotheses ?*

<details>

<summary>Working hypotheses about our question</summary>

H0 = **The average heart rate is not different or lower in patients with heart disease than in other patients**.

H1 = **The average heart rate is higher in patients with heart disease**.

</details>

In our question, and therefore in our assumptions, there is the notion of **“higher".** The biological question is therefore **directional**. So we will be in a situation where we will be on a **one-tailed test**.

## Independent or paired ?

This is an important point when doing **statistics**. It is important to know whether our data is **independent** or **paired**.

-   *Independent* : The choice of elements in the first sample has no influence on the choice of elements in the second sample.

-   *Paired* : The choice of elements in the first sample influences the choice of elements in the second sample.

*What would it be in our case ?*

<details>

<summary>Independent or paired data ?</summary>

All patients in our dataset are **different** from each other and were not selected based on others, so we can say that our samples are **independent**.

</details>

## Test choice

There are many statistical tests available, and it is easy to get confused. Depending on the **biological question** being asked and also on the **data itself**, one test will be used rather than another.

Based on our question, we will seek to **compare groups**, in this case one diagnostic group (**heart disease**) versus the **other diagnostic groups**. So we will compare two groups.

Now the question we must ask ourselves is : **Are we trying to compare averages, distributions, proportions, etc.?**

### Compare means

To compare means, there is a determination key for choosing the right test based on the data. There are two keys depending on whether the samples are independent or paired.

![](./Images/determination_ind_samples.drawio.svg){fig-align="center"}

![](./Images/determination_paired_samples.drawio.svg){fig-align="center"}

Based on those information, what should we use as a determination key ?

<details>

<summary>Response</summary>

As we identified above, we are working with **independent samples**. The aim is to compare the **average heart rates** between **patients with heart disease** and **other patients** (2 group). We will start with **the key for independent samples**.

</details>

### Conditions of application

The first **condition of application** that we seek to verify is **normality**. To check normality, the first thing we can do is look at how the data we are interested in is **distributed**, graphically.

#### Histogram

Here, we will look at the distribution of heart rates in patients with heart disease.

```{r, eval = F}
hist(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"],
     main = "Heart rate from patients with heart disease",
     col = "red",
     xlab = "Heart rate")
```

```{r, echo=F}
hist(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"],
     main = "Heart rate from patients with heart disease",
     col = "red",
     xlab = "Heart rate")
```

How can this histogram be interpreted?

<details>

<summary>Response</summary>

You have to look at the **shape of the distribution**. If the data approximates a **normal distribution**, then you will see **symmetry** in the data with a **bell shape**.\
In our case, there seems to be a symmetry emerging with this bell shape : **we can therefore say that the data appear to follow a normal distribution**.

</details>

#### Q-Q plot

A **quantile-quantile** (Q-Q) plot is used to check whether a data set follows a **theoretical distribution**. We do not **compare the data directly**, but rather **their position in the distribution** that we observe.

For example, in our case, we will classify the **heart rate values for patients with heart disease in ascending order** and associate each value with **the corresponding quantile** (10%, 25%, 50%, etc.). For this same **quantile**, we look at what it corresponds to in the **theoretical distribution** (normal distribution in our case).

To be more specific: **the median corresponds to the 50% quantile, which corresponds to 0 in a normal distribution**.

The closer the observed quantiles are to the theoretical quantiles, the closer our data is to a normal distribution.

```{r, eval=F}
# Q-Q plot with qqnorm function (equivalent of qqplot but for normality)
qqnorm(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
qqline(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
```

```{r, echo=F}
qqnorm(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
qqline(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
```

What can be said about this plot?

<details>

<summary>Response</summary>

Despite **slight discrepancies** in the most extreme values, the **observed quantiles appear to follow the theoretical quantiles**. This allows us to hypothesize that **the heart rate values of patients with heart disease appear to follow a normal distribution**.

</details>

#### Statistical approach to check normality

To check for statistical normality, there is a test called the **Shapiro-Wilk test**.

This tool allows you to test whether a **distribution deviates significantly from a normal distribution**. It is the element that completes the graphical inspection.

As with any statistical analysis, we start with an **initial hypothesis** that we seek to verify through testing.

-   **H0** : the data **follow a normal distribution**
-   **H1** : the data **do not follow a normal distribution**

If the p-value obtained via the Shapiro test is less than 0.05, then we can **reject the null hypothesis H0**.

NOTE: **Rejecting H0 means that we detect a statistical difference between our data and the Normal distribution**.

```{r, eval=F}
# Test normality for patients with heart disease
shapiro.test(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
# Test normality for others patients
shapiro.test(clinical_data$heart_rate[clinical_data$diagnosis != "Heart Disease"])
```

**Shapiro-Wilk test on the heart rate variable for patients with heart disease :**

```{r, echo=F}
shapiro.test(clinical_data$heart_rate[clinical_data$diagnosis == "Heart Disease"])
```

**Shapiro-Wilk test on the heart rate variable for patients with other diagnosis :**

```{r, echo=F}
shapiro.test(clinical_data$heart_rate[clinical_data$diagnosis != "Heart Disease"])
```

How can we interpret those results ?

<details>

<summary>Response</summary>

We consider that we set a standard **threshold of 5%** (i.e., 0.05). Here, both tests give **p-values that are greater than 0.05**. This means that we **do not detect any significant deviation from normality**. In this case, we can move on to a **parametric test, i.e., Student's t-test**.

</details>

### Homogeneity of variances

This is the **second condition of application** that we need.

When **comparing means between groups**, it is assumed that the **variability** of the measurements is **comparable** in each group.
This means that the **differences** observed are **not due to one group** being **much more variable than the other**.

The first check we can do is graphically, with **boxplots** like we did earlier.

```{r, eval=F}
ggplot(data = clinical_data, aes(x=diagnosis, y = heart_rate)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

```{r, echo=F}
ggplot(data = clinical_data, aes(x=diagnosis, y = heart_rate)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

Thanks to this simple representation, we can look at the **appearance of the boxplots**. 

- Are the sizes comparable? 
- Are there any outliers? 
- Are the whiskers similar in size or very different?

What do you think about variability here ?

<details>

<summary>Response</summary>

What we observe here is that, overall, **the box and whisker sizes are fairly similar**. However, it should be noted that for hypertension, the box and whiskers are slightly smaller.
There are **no major differences**.

</details>

For the homogeneity of variances, we can check statistically too with a **Fisher test**.
This test is used **to determine whether two groups exhibit comparable variability**. 
**Are the measurements more scattered in one group than in the other ?**

We then make the following assumption :

- **H0** : the variances of **the two groups are equal**
- **H1** : the variances are **different**

If the p-value obtained is less than 0.05, then we can reject **the null hypothesis H0**.

## Interpretation of the results



# Study 2 : une analyse multi-groupes pour anova / krukall
