---
title: "Practical stats"
format: 
  html: 
    code-fold: true
    embed-resources: true 
editor: visual
---

```{=html}
<style>
details > summary {
  padding: 4px;
  width: 400px;
  background-color: #eeeeee;
  border: none;
  box-shadow: 1px 1px 2px #bbbbbb;
  cursor: pointer;
}

details > p {
  background-color: #E0F8EC;
  padding: 4px;
  margin: 0;
  box-shadow: 1px 1px 2px #bbbbbb;
}
header h1 {
      font-size: 70px;
      font-weight: 600;
      font-palette: tomato;
    }
</style>
```

```{r, setup, echo=F, warning=FALSE,message=FALSE}
require("knitr")
library(readr)
library(tidyverse)
clinical_data <- read_csv("Datasets/clinical_data.csv")
```
# Objectives 

The purpose of this section is to provide you with the basics needed to analyze biological data using simple statistics.
The main objective is to provide you with the tools to :

-   Choose a statistical method appropriate for a biological question
-   Interpret the results correctly
-   Analyse data with R

# Discover and explore data : `clinical_data`

## Load the data 

First step to do is to load our dataset called `clinical_data` in our R environment.

comment ?

## Description of the data

rappel types de variables dans le dataset (descriptive part)
rappel head et str 

summary

# Study 1 : *Do patients with heart disease have a higher heart rate ?*

## First visualisation

```{r, echo=F}
ggplot(data = clinical_data, aes(x=diagnosis, y = heart_rate)) +
  geom_boxplot() + 
  labs(x="Diagnosis", y = "Heart rate") +
  theme_bw()
```

## Hypotheses

H0 et H1
unilateral (one-tailed) et pourquoi ? diff avec bilateral (two-tailed)

## Independent or paired ?

## Test the parameters

## Application of the statistical test

## Interpretation of the results


# Study 2 : une analyse multi-groupes pour anova / krukall